---
title: "DSC 520 Final Project Step 2"
author: "Taylor Imhof"
date: "11/11/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# libraries
library(DataExplorer)
library(dplyr)
library(Hmisc)
library(ggplot2)
library(purrr)
library(data.table)
```

## Data Import and Cleaning Efforts
Fortunately, all of my data sets came in .csv format, so reading in the data to
data frames was straightforward using the built-in read.csv() function. 

```{r, echo=T, results='hide'}
# read in data to data frames
happiness_2018_df <- read.csv('data/happiness_2018.csv', stringsAsFactors = FALSE)
lifestyle_df <- read.csv('data/lifestyle.csv', stringsAsFactors = FALSE)
human_df <- read.csv('data/human_development.csv', stringsAsFactors = FALSE)
```


To get a basic understanding of how the data is designed, I ran str() on each of
them to see the different data types.
```{r, echo=T, results='hide'}
# check structure of each data frame
str(happiness_2018_df)
str(lifestyle_df)
str(human_df)
```

I did not really like how the original column names were written, so I made some
slight adjustments to make it easier for me to understand and code later on.

```{r, include=T, results='hide'}
# rename columns for easier coding
old1 <- c('Overall.rank', 'Country.or.region', 'GDP.per.capita', 'Social.support',
          'Healthy.life.expectancy', 'Freedom.to.make.life.choices', 'Perceptions.of.corruption')
new1 <- c('Rank', 'Country', 'GDP', 'socialSupport', 'lifeExpectancy', 'choiceFreedom',
          'Corruption')
old2 <- c('HDI.Rank', 'Human.Development.Index..HDI.', 'Life.Expectancy.at.Birth',
         'Expected.Years.of.Education' ,'Mean.Years.of.Education', 
         'Gross.National.Income..GNI..per.Capita','GNI.per.Capita.Rank.Minus.HDI.Rank')
new2 <- c('HDIrank', 'HDI', 'lifeExpectancy', 'expectedEducationYears', 
          'meanEducationYears', 'GNI', 'GNIminusHDIrank')
setnames(happiness_2018_df, old=old1, new=new1)
setnames(human_df, old=old2, new=new2)
```

Right off the bat, I noticed that the `Corruption` column had chr values. To fix
this, I first set the imputed all of the NA values with the mean of the 
distribution for that column. Then, I converted the column's values to numeric 
values using the as.numeric() function.

```{r, echo=T, results='hide'}
# convert perception.of.corruption from chr to num
happiness_2018_df$Corruption[is.na(happiness_2018_df$Corruption)] <- mean(happiness_2018_df$Corruption, na.rm=TRUE)
happiness_2018_df$Corruption <- as.numeric(as.character(happiness_2018_df$Corruption))
str(happiness_2018_df)
```

In a similar vein, the `DAILY_STRESS` column in the lifestyle data frame was also
interpreted as characters. I simply followed the same logic I used above to
correct for this.

```{r, echo=T, results='hide'}
lifestyle_df$DAILY_STRESS[is.na(lifestyle_df$DAILY_STRESS)] <- mean(lifestyle_df$DAILY_STRESS, na.rm=TRUE)
lifestyle_df$DAILY_STRESS <- as.numeric(as.character(lifestyle_df$DAILY_STRESS))
str(lifestyle_df)
```

Next, in the lifestyle data frame, the column `GENDER` used character values to 
denote gender, so I went ahead and converted these to binary values so that they
could be better used during analysis. I created a new column `SEX` and converted 
all Male values to 1 and all female values to 0.

```{r, include=T, results='hide'}
# convert GENDER column to binary values
lifestyle_df$SEX[lifestyle_df$GENDER=='Male'] <- 1
lifestyle_df$SEX[lifestyle_df$GENDER=='Female'] <- 0
str(lifestyle_df)
```

Since the lifestyle data frame doesn't have any values that can be used to merge
with the other data frames, I decided to use it to analyze which lifestyle factors
can improve overall health (this data set used a calculated work-life-balance).
The other two data frames both contained country names, so merging them together
was straightforward. Using the merge() function, I was able to combine the two 
together.

```{r, include=TRUE, results='hide'}
# merge happiness and HDI dfs together on country name
combo_df <- happiness_2018_df %>% merge(human_df, by='Country', all.x = T)
str(combo_df)
```

In this new data frame, the `GNI` column contained character values. I predicted
this column would be a useful predictor, so I need to convert it to numeric values.
Following a similar procedure I used earlier, I converted all the values using
the as.numeric() function. A pesky issue that was giving me grief was the starting
values all contained commas. To remedy this, I used gsub() to replace them.

```{r, include=TRUE, results='hide'}
# replace commas with white space
combo_df$GNI <- as.numeric(gsub(",", "", combo_df$GNI))
combo_df$GNI[is.na(combo_df$GNI)] <- mean(combo_df$GNI, na.rm=TRUE)
combo_df$GNI <- as.numeric(combo_df$GNI)
```

The next thing I decided to do was make sure that my data did not contain any
NA values. I found a pretty slick implementation of the map_df from the purrr
package that runs through each column and sums up all the NA values. There were
quite a few NA values for columns that were added during the merge. This was 
expected, as there were likely country's not included in both. To remedy this,
I opted to impute the means for these NA values. Lastly, I did the same operation
to check for NA values on the lifestyle data frame. However, fortunately all of
the values contained in this data set were "clean" and contained no NA values.

```{r, include=TRUE, results='hide'}
# check for NA values using map_df from purrr
combo_df %>% map_df(~sum(is.na(.)))

# impute mean for columns with NA values
for(i in 1:ncol(combo_df)){
  combo_df[, i][is.na(combo_df[,i])] <- mean(combo_df[,i], na.rm=TRUE)
}

# check again for NA values to ensure imputation was successful
combo_df %>% map_df(~sum(is.na(.)))

# check for NA values in lifestyle_df
lifestyle_df %>% map_df(~sum(is.na(.)))

# impute mean for columns with NA values
for(i in 1:ncol(lifestyle_df)){
  lifestyle_df[, i][is.na(lifestyle_df[,i])] <- mean(lifestyle_df[,i], na.rm=TRUE)
}

# convert lifestyle values from integer to numeric
lifestyle_df[1:22] <- lapply(lifestyle_df[1:22], as.numeric)
str(lifestyle_df)
```

The last thing I decided to do during the data cleaning stage was narrow down
columns for each data frame. During the merge of the HDI and WHR data sets, two
life expectancy columns were created. However, only one seems to use years as a
metric, so I opted to keep this column. There are two rank scores as well, both
of which I decided would not be useful for analysis so they were dropped. In the
lifestyle data frame, I decided the timestamp info would not be useful. Also,
since I created a new variable of binary values based on the `GENDER` column, it
was dropped. The age column also contained some strange values, so I thought it
would be easier to just leave it out of the final data frame.

```{r, include=T, results='hide'}
#
dropCols1 <- c('Rank','lifeExpectancy.x','HDIrank', 'GNIminusHDIrank')
combo_df <- combo_df[,!(names(combo_df) %in% dropCols1)]

dropCols2 <- c('ï..Timestamp', 'AGE', 'GENDER')
lifestyle_df <- lifestyle_df[,!(names(lifestyle_df) %in% dropCols2)]
str(lifestyle_df)
```


## Final Two "Cleaned" Data Frames For Analysis
After performing the munging operations outlined above, I was pretty happy with
my two data sets. Below is the output of the str() function on both of them for
review.
```{r, echo=FALSE}
str(combo_df)
str(lifestyle_df)

```


## Plots and Data Transformations
For my two final data frames, I decided that `Score` and `WORK_LIFE_BALANCE_SCORE`
would be the best target variables. I ran a quick hist() on both of these features
to ensure that they were both normally distributed. Both appeared to be normal
distributions. In fact, the `WORK_LIFE_BALANCE_SCORE` looked like a textbook
example of the normal curve.

```{r, include=TRUE}
par(mar=c(1,1,1,1))

# check distribution of target variables for both data frames
hist(combo_df$Score)
hist(lifestyle_df$WORK_LIFE_BALANCE_SCORE)
```

Next, to find which features would make for good predictors, I ran a Pearson's
correlation test on the numeric values of each data frame. Some of the features
that stuck out for the combo_df were `GDP`, `HDI`, `socialSupport`, and
`lifeExpectancy`. After looking at the distributions for each of these potential
explanatory features, there were two negatively skewed distributions that could
use transformation. These were the `socialSupport` and `lifeExpectancy` columns
respectively. After creating a new variable contained the squared values,
the distributions did appear to take on a more normal distribution.

```{r, include=FALSE, warning=FALSE}
par(mar=c(1,1,1,1))

# run cor test to identify potential independent variables
cor(combo_df[-1], method='pearson')
```

```{r, include=TRUE}
# check distributions for normality
ggplot(combo_df, aes(x=GDP)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white') +
  geom_density(alpha=.2, fill='#FF6666')

ggplot(combo_df, aes(x=HDI)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white') +
  geom_density(alpha=.2, fill='#FF6666')

ggplot(combo_df, aes(x=socialSupport)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white') +
  geom_density(alpha=.2, fill='#FF6666')

ggplot(combo_df, aes(x=lifeExpectancy.y)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white') +
  geom_density(alpha=.2, fill='#FF6666')

# square values of life expectancy for more normal distro
combo_df$squareLifeExpectancy <- combo_df$lifeExpectancy.y ^2
ggplot(combo_df, aes(x=squareLifeExpectancy)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white') +
  geom_density(alpha=.2, fill='#FF6666')

# square values for social support for normal distro
combo_df$squareSocialSupport <- combo_df$socialSupport ^2
ggplot(combo_df, aes(x=squareSocialSupport)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white') +
  geom_density(alpha=.2, fill='#FF6666')
```

For the lifestyle data frame with the work-life balance score as the dependent
variable, there did not appear to be any variables with strong correlations. The
ones of note were `PLACES_VISITED`, `TIME_FOR_PASSION`, `SUPPORTING_OTHERS`, and
`ACHIEVEMENT`.

```{r, include=FALSE, warning=FALSE}
par(mar=c(1,1,1,1))

cor(lifestyle_df, method='pearson')
```

```{r, include=TRUE}
ggplot(lifestyle_df, aes(x=PLACES_VISITED)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white', bins=10) +
  geom_density(alpha=.2, fill='#FF6666')

ggplot(lifestyle_df, aes(x=ACHIEVEMENT)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white', bins=10) +
  geom_density(alpha=.2, fill='#FF6666')

ggplot(lifestyle_df, aes(x=SUPPORTING_OTHERS)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white', bins=10) +
  geom_density(alpha=.2, fill='#FF6666')

# take square root of time for passion to get more normal distro
lifestyle_df$squarePassionTime <- lifestyle_df$TIME_FOR_PASSION ^(1/2)

ggplot(lifestyle_df, aes(x=TIME_FOR_PASSION)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white', bins=10) +
  geom_density(alpha=.2, fill='#FF6666')
ggplot(lifestyle_df, aes(x=squarePassionTime)) + 
  geom_histogram(aes(y=..density..), colour='black', fill='white', bins=10) +
  geom_density(alpha=.2, fill='#FF6666')

```

## Scatter Plots and Trend Analysis

After selecting a few potential explanatory variables for each data set, I
decided to use scatter plots to see if any linear relationships could be spotted.
I used ggplot() functionality and added a regression line using the additional
geom_smooth() function. As I had predicted, all four of my selected variables
appeared to be positively related with the target variables.


```{r, include=TRUE, warning=FALSE}
par(mar=c(1,1,1,1))

ggplot(combo_df, aes(x=squareLifeExpectancy, y=Score)) + 
  geom_point() +
  geom_smooth(method=lm)

ggplot(combo_df, aes(x=GDP, y=Score)) + 
  geom_point() +
  geom_smooth(method=lm)

ggplot(combo_df, aes(x=squareSocialSupport, y=Score)) + 
  geom_point() +
  geom_smooth(method=lm)

ggplot(combo_df, aes(x=HDI, y=Score)) + 
  geom_point() +
  geom_smooth(method=lm)
```

```{r, echo=TRUE, warning=FALSE}
par(mar=c(1,1,1,1))

ggplot(lifestyle_df, aes(x=PLACES_VISITED, y=WORK_LIFE_BALANCE_SCORE)) + 
  geom_point() +
  geom_smooth(method=lm)
ggplot(lifestyle_df, aes(x=ACHIEVEMENT, y=WORK_LIFE_BALANCE_SCORE)) + 
  geom_point() +
  geom_smooth(method=lm)
ggplot(lifestyle_df, aes(x=SUPPORTING_OTHERS, y=WORK_LIFE_BALANCE_SCORE)) + 
  geom_point() +
  geom_smooth(method=lm)
ggplot(lifestyle_df, aes(x=squarePassionTime, y=WORK_LIFE_BALANCE_SCORE)) + 
  geom_point() +
  geom_smooth(method=lm)
```

## Interpreting the Results
After conducting initial EDA efforts, including data importation, cleaning, and
transformation, there are a quite a few observations that can be made. First,
it is important to note that a lot of these are relatively intuitive (e.g., people
who live in richer countries tend to be 'happier'). However, it was very useful
to see how to go about supporting these statements with hard numbers and statistics.

One of the more interesting conclusions that can be made from the results is that
countries with higher life expectancies tend to have a happier populace. I believe
this is strongly supported with the additional trend that countries with a stronger
sense of social support (i.e., social programs) also tend to have a happier
population. Based on these findings, if I had the ear of my country's policymakers,
I would ensure they understood these implications. Perhaps spending more on developing
robust social programs will lead to healthier and happier people and a more 
productive society overall.

Moreover, summarizing the results from analyzing the lifestyle data set, there are
a few prescriptions I would give if I were say, giving a speech about how to lead
a happier life. The four predictors that I focused in on were again the number of
places visited, sense of achievement, time spent on passions, and supporting others.
The first makes intuitive sense; people who travel to more places are likely to be
more financially secure than those who do not travel as often. However, I could
compound this with the findings from the previous data set to encourage policymakers 
further to develop social programs where people can travel abroad for service 
projects.

Another insight that I found most fascinating was the positive link between
happiness and support for other people. Understanding that helping individuals in
one's community is powerful. With the depression epidemic that is claiming so
many lives to suicide, perhaps a new method of treating such ailments would be
strong encouragement of participation in local service projects.


```{r, include=FALSE, eval=FALSE}
# useful for pop-out window of each data frame
utils::View(happiness_2018_df)
utils::View(lifestyle_df)
utils::View(human_df)
utils::View(combo_df)
```